---
title: "Case study 1 - How does a bike-share navigate speedy success?"
author: "Donovan Rodriguez"
date: "2023-04-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercise Full Year Analysis

This analysis is the case study "'How does a bike-share navigate speedy success?’ (found here: https://divvy-tripdata.s3.amazonaws.com/index.html). The purpose of this analysis is to practice the skills learned in the course.

```{r}
library(tidyverse)
library(lubridate)
```
```{r}
getwd()
setwd("/Users/donovanrodriguez/Documents/Database/CSV")
```

in the time that i clean the data in excel i saw tthat trhough the time the casual member was increase in the pass of time and in the most actual data the casual members now are our main source 

### Step 1 Collected Data
```{r}
J01 <- read_csv("/Users/donovanrodriguez/Documents/Database/CSV/202201-divvy-tripdata.csv")

F02 <- read.csv("/Users/donovanrodriguez/Documents/Database/CSV/202202-divvy-tripdata.csv")

M03 <- read_csv("/Users/donovanrodriguez/Documents/Database/CSV/202203-divvy-tripdata.csv")

A04 <- read.csv("/Users/donovanrodriguez/Documents/Database/CSV/202204-divvy-tripdata.csv")

M05 <- read_csv("/Users/donovanrodriguez/Documents/Database/CSV/202205-divvy-tripdata.csv")

J06 <- read_csv("/Users/donovanrodriguez/Documents/Database/CSV/202206-divvy-tripdata.csv")

J07 <- read_csv("/Users/donovanrodriguez/Documents/Database/CSV/202207-divvy-tripdata.csv")

A08 <- read_csv("/Users/donovanrodriguez/Documents/Database/CSV/202208-divvy-tripdata.csv")

S09 <- read_csv("/Users/donovanrodriguez/Documents/Database/CSV/202209-divvy-publictripdata.csv")

O10 <- read_csv("/Users/donovanrodriguez/Documents/Database/CSV/202210-divvy-tripdata.csv")

N11 <- read_csv("/Users/donovanrodriguez/Documents/Database/CSV/202211-divvy-tripdata.csv")

D12 <- read_csv("/Users/donovanrodriguez/Documents/Database/CSV/202212-divvy-tripdata.csv")
```

### Step 2 Wrangle data and combine into a single file

```{r}
colnames(J01)
colnames(F02)
colnames(M03)
colnames(A04)
colnames(M05)
colnames(J06)
colnames(J07)
colnames(A08)
colnames(S09)
colnames(O10)
colnames(N11)
colnames(D12)
```

```{r}
J01 <-  mutate(J01, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
                   
F02 <-  mutate(F02, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type))

M03 <-  mutate(M03, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type))

A04 <-  mutate(A04, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type))

M05 <-  mutate(M05, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type))

J06 <-  mutate(J06, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 

J07 <-  mutate(J07, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 

A08 <-  mutate(A08, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 

S09 <-  mutate(S09, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 

O10 <-  mutate(O10, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 

N11 <-  mutate(N11, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 

D12 <-  mutate(D12, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
```

```{r}
F02$started_at <- ymd_hms(F02$started_at, tz = "UTC")
F02$ended_at <- ymd_hms(F02$ended_at, tz = "UTC")
M03$started_at <- ymd_hms(M03$started_at, tz = "UTC")
M03$ended_at <- ymd_hms(M03$ended_at, tz = "UTC")
A04$started_at <- ymd_hms(A04$started_at, tz = "UTC")
A04$ended_at <- ymd_hms(A04$ended_at, tz = "UTC")
```
#### Stack individual quarter's data frames into one big data frame

```{r}
all_trips <- bind_rows(J01, F02, M03, A04, M05, J06, J07, A08, S09, O10, N11, D12)
```

### Step 3 Clean up and add data to prepare for analysis

#### Inspect the new table that has been created

```{r}
colnames(all_trips)  #List of column names
nrow(all_trips)  #How many rows are in data frame?
dim(all_trips)  #Dimensions of the data frame?
head(all_trips)  #See the first 6 rows of data frame.
tail(all_trips) # See the last 6 rows of data frame.
str(all_trips)  #See list of columns and data types (numeric, character, etc)
summary(all_trips)  #Statistical summary of data. Mainly for numerics
```

```{r}
# Reassign to the desired values
all_trips <-  all_trips %>% 
  mutate(member_casual = recode(member_casual
                           ,"Subscriber" = "member"
                           ,"Customer" = "casual"))

```


```{r}
table(all_trips$member_casual)
```


```{r}
all_trips$date <- as.Date(all_trips$started_at) #The default format is yyyy-mm-dd
```

```{r}
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```

#### Add a "ride_length" calculation to all_trips (in seconds)

```{r}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
```

# Inspect the structure of the columns
```{r}
str(all_trips)
```

#### Convert "ride_length" from Factor to numeric so we can run calculations on the data

```{r}
is.factor(all_trips$ride_length)
class(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
class(all_trips$ride_length)
is.numeric(all_trips$ride_length)
```

#### Remove "bad" data, the dataframe includes a few hundred entries when bikes were taken out of docks and checked for quality by Divvy or ride_length was negative, we will create a new version of the dataframe (v2) since data is being removed

```{r}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]

```


### Step 4 Conduct descriptive analysis

#### Descriptive analysis on ride_length (all figures in seconds)

```{r}
summary(all_trips_v2$ride_length)

mean(all_trips_v2$ride_length, na.rm = TRUE) #straight average (total ride length / rides)

median(all_trips_v2$ride_length, na.rm = TRUE) #midpoint number in the ascending array of ride lengths

max(all_trips_v2$ride_length, na.rm =TRUE) #longest ride

min(all_trips_v2$ride_length, na.rm = TRUE) #shortest ride

```
```{r}
# Compare members and casual users
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)

```

```{r}
# See the average ride time by each day for members vs casual users
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

```{r}
# Notice that the days of the week are out of order. Let's fix that.
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

```{r}
# Now, let's run the average ride time by each day for members vs casual users
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)

```

```{r}
# analyze ridership data by type and weekday
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%  #groups by usertype and weekday
  summarise(number_of_rides = n()							#calculates the number of rides and average duration 
  ,average_duration = mean(ride_length)) %>% 		# calculates the average duration
  arrange(member_casual, weekday)								# sorts
```

```{r}
# Let's visualize the number of rides by rider type
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")

```

```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  filter(!is.na(weekday)) %>% # Remove rows where weekday is NA
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")

```


As you can see in the visualization the casual member use more in weekends and the members use more in weekday this can be useful we can use an ad for weekdays for workers and in weekends we can use an ad for casual to have a ride in the park or almost here in mexico city we have days that the avenue Reforma is closed and you can use the bicycle.
```{r}
# Let's create a visualization for average duration
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%
  filter(!is.na(weekday)) %>%
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")

```

### Step 5 Export summary file for further analysis
#### Create a csv file that we will visualize in Excel, Tableau, or my presentation software

```{r}
counts <- aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

```{r}
write.csv(counts, file = '/Users/donovanrodriguez/Documents/Database/avg_ride_length.csv')
```

To see the visualization you can go to my Tableu here: https://public.tableau.com/app/profile/donovan7710
